---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    keep_md: yes
---

## Source for the Data
Data for this document can be found here: [Activity monitoring dataset](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip)  
Alternatively, a zipfile `activity.zip` containing the data is present in the repository itself.  

## Loading and preprocessing the data
Load the data from the csv file. Change the file path to where the data file is 
located.  

```{r Reading Data}
filePath <- "../" # change this to to where the data is located on your system
fullPath <- paste("../", "activity.csv", sep = "")
rawData <- read.csv(fullPath, header = TRUE, stringsAsFactors = FALSE)

## convert the date column into the right format
rawData$date <- as.Date(rawData$date) 

## create dataset removing NA values
cleanData <- subset(rawData, is.na(rawData$steps) == FALSE)
```

## What is mean total number of steps taken per day?
First, we calculate the total number of steps taken per day and plot its 
histogram.  

```{r First question}
## create vector of steps per day using dates information
options(scipen=999)

## using date as a factor to sum total steps (per day)
stepPerDay <- tapply(X = cleanData$steps,INDEX = cleanData$date, FUN = sum)

## calculating mean and median values
meanSteps <- round(mean(stepPerDay),digits = 2)
medianSteps <- median(stepPerDay)
```
```{r Fig1 Histogram of Clean data}
## plot the histogram of total number of steps taken per day
hist(stepPerDay,xlab = "Steps per day", 
     main = "Figure 1: Histogram of total number of steps 
     taken each day (Cleaned Data)", breaks = 10, 
     col = "grey",ylim = c(0,25))
```


The data for total number of steps per day has a mean of **`r meanSteps`** 
and a median of **`r medianSteps`**.


## What is the average daily activity pattern?
First, we calculate the average number of steps (across all days) taken within a time interval.

```{r Second question}
## calculating the average steps for each interval (across dates), using steps as
## a factor
meanByInterval <- tapply(cleanData$steps,INDEX = cleanData$interval, FUN = mean)
intervals <- unique(cleanData$interval)
```
```{r Fig3 Avg Daily Activity Pattern}
## constructing a time series plot to denote the activity (steps) on an 'average' day
plot(x = intervals, y = meanByInterval, type = 'l', 
     xlab = "Time Intervals (minutes)", ylab = "Mean Steps (across days)", main = "Figure 2: Average Daily activity pattern", col = "blue3")

## calculating the time interval with max average steps
maxStepsInterval <- intervals[which.max(meanByInterval)]
```

The 5-minute interval, on average across all the days in the dataset, that contains the maximum number of steps is the **`r maxStepsInterval`th** interval.

## Imputing missing values

The dataset contains several NA values. In order to impute these values, we first calculate the total number of rows with NA values for steps.

```{r Calculating NA rows}
naRows <- sum(is.na(rawData$steps))
```

There are **`r naRows` rows** that contain NA values for steps.  
To impute these values, we 

 * Use the **means** for each time `interval` across all dates and  
 * Replace the NA values depending on the time interval they refer too

```{r Imputing Values}
intervalMeanValues <- as.data.frame(cbind(intervals, meanByInterval))

imputedData <- rawData
for (i in 1:nrow(imputedData)){
    if (is.na(imputedData[i,1]) == T){
        currentInterval <- imputedData[i,3]
        ## Replace NA value with the mean calculated for the interval
        imputedData[i,1]<- intervalMeanValues[intervals == currentInterval, 2]
    }
}
```
```{r Fig3 Histogram of Imputed Data}
## plotting a histogram of the total number of steps taken each day
impstepPerDay <- tapply(X = imputedData$steps,INDEX = imputedData$date, FUN = sum)
impmeanSteps <- round(mean(impstepPerDay),digits = 2)
impmedianSteps <- round(median(impstepPerDay), digits = 2)
## plot the histogram of total number of steps taken per day
hist(impstepPerDay,xlab = "Steps per day", 
     main = "Figure 3: Histogram of total number of steps
     taken each day (Imputed Data)", breaks = 10, 
     col = "grey", ylim = c(0,25))

```

Based on the imputed dataset, the histogram for total number of steps per day has a mean of **`r impmeanSteps`** and a median of **`r impmedianSteps`**.

If we compare _Figure 1_ and _Figure 3_, we can see that the overall shape of the histogram has been preserved, with more data being added in _Figure 3_.

### Difference from results obtained while excluding missing values

Comparing the above results with the mean of **`r meanSteps`** and median of **`r medianSteps`** obtained with the dataset excluding NA values we see that:

1. **Mean** of total number of steps per day remains the same
2. **Median** of toal number of steps per day is slightly higher for the imputed data than the clean data. It is also not a whole number (as one would expect steps to be) and is equal to the **Mean**

### Understanding the impact of imputing missing data on estimates

This change in the value of the **Median** is primarily because of our choice to impute with the **mean values** for each interval (across dates).  
As a result of **`r naRows`** NA values being replaced with the mean value, the _"likelihood of intervals having the mean value"_ has increased. This has caused the median to shift in value towards the mean of the dataset.


## Are there differences in activity patterns between weekdays and weekends?

In order to split our imputed data into weekdays and weekends, we add another variable `weekDay` to our dataset.

```{r Fig4 Weekday and Weekend Activity pattern, fig.width = 10}

## Add new column to data.frame calculating the day of week based on dates
imputedData$weekDay<- weekdays(imputedData$date)

## Replace days with weekday or weekend and converting the variable to a factor
for (i in 1: nrow(imputedData)){
    if (imputedData[i,4] == "Sunday" || imputedData[i,4] == "Saturday"){
        imputedData[i,4] <- "weekend"
    }
    else {
        imputedData[i,4]<- "weekday"
    }
}
imputedData$weekDay <- as.factor(imputedData$weekDay)

## Split the mean data by interval and by weekday/ weekend type
splitByDay <- tapply(imputedData$steps,list(imputedData$interval,imputedData$weekDay),FUN = mean)
splitByDay <- as.data.frame(splitByDay)

## Configure line plot panels to compare activity pattern on weekdays and weekends
par(mfrow = c(1,2), mar = c(4,4,2,1), oma = c(0,0,2,0))

## Setting ylim for easy comparison
plot(x = as.numeric(row.names(splitByDay)),y = splitByDay$weekday,type = "l",
     xlab = "Time Intervals (minutes)", ylab = "Mean Steps (across days)", 
     main = "Weekdays", col = "blue3", ylim = c(0,250))
     
plot(x = as.numeric(row.names(splitByDay)),y = splitByDay$weekend,type = "l",
     xlab = "Time Intervals (minutes)", ylab = "", main = "Weekends", 
     col = "cyan4", ylim = c(0,250))

mtext("Figure 4: Average Daily Activity Pattern",outer = TRUE)

weekdaySum <- round(sum(splitByDay$weekday))
weekendSum <- round(sum(splitByDay$weekend))
```


On inspecting _Figure 4_, we see that:

1. Overall daily activity has shifted towards the later intervals during weekends as compared to weekdays.
2. Total steps on an Average weekday is approx. **`r weekdaySum`** while Total steps on an Average weekend is approx. **`r weekendSum`**.


